Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-05 paper.qmd
most final for words...still need to fix pipeline

@@

<!-- teaser opening --><!-- might need a rewrite - first jotting down -->
The quest for space is a driver for technological development.
Key milestones include the first human traval to space and surrounding the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
These developments helped to identify options for the wider use of space, benefits, and its commercialisation.
This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable prooduction of goods in a non-gravity environment, services, and recently leisure.
The increased number of launches, however, also increased the number of artefacts <!-- falling down and --> re-entering the atmosphere.

@@

<!-- APPROACH / METHOD --------------------------------------------------- -->
<!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->
<!-- as a minimum, we should say something about the WHAT -->
In general, the method for estimating the risk to commercial aircraft from reentering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
In that respect, a robust determination of flight density is a key aspect.

@@

This work focusses on the European region.
The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1].
To decrease spatial complexity, the data is binned using Uber's H3 geospatial indexing system [@uber-h3]. 
The reduction of the spatial complexity enables fast processing.
For this work, we employ a H3-cell-resolution of 3. 
This corresponds to hexagons of an average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

@@

H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded polygon).

@@

[^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's 42 Member States and 2 Comprehensive Member States covering almost all \acr{ECAC} States.

<!-- next sentence feels like an orphan ... do we need this here? ------>
Next sections will develop the mathematical background for the evaluation of collision and casualties risk of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
The chart shows the expected outcomes in terms of major traffic flows, and areas of high or low traffic.
For example, cells with a density of zero are clearly visible for areas of war or no-fly zones
(Ukraine, Libia or Syria). The major traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted showing densities of more than 10.

@@

```{r density-winter-res-3}
#| out-width: "110%"
#| echo: false
knitr::include_graphics("media/figures/flight_density_2024-07-05_3.png")
```

Flight density in the EUROCONTROL Area on the 5th of July 2024, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$).
:::

@@

Safety and collision risk awareness is directly linked to the probability of a re-entering space object colliding with a transversing flight.
This probability, $p_k$, of a debris (point) object $k$, impacting an
aircraft (point) object over the area, $a$, is expressed as:

@@

where 
* $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
* $\sigma(\vec{x})$ is the impact probability density of a debris object $k$
in the position $\vec{x}$, and 
* $\varepsilon_k$ is the effective exposed area, i.e., the effective risk area for the $k^{th}$ debris object impacting
the aircraft (see @sec-eea later on).

Our approach builds on the discretisation of the geo-spatial risk area.
For this purpose we use a hexagon-based tesselation pattern.
Uber's H3 method ensures non-overlapping hexagons, with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
Using H3 to discretise the area of interest $a$ gives

@@

The impact probability $p_{k}(h)$ for each cell $h$ is determined by the following

$$
p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h).
$$

@@

For the actual collision encounter between a re-entering object and an aircraft, we assume the aircraft impact area to be much larger than the debris object area.
Thus, the area spanned by the debris object is smaller than the aircraft dimensions. 
This assumption holds for practically all commercial aircraft within the scope of this study, and we can neglect the debris impact area.
Accordingly, we can also simplify the geometry of the aircraft and model it as a rectangular solid with dimesions of wingspan ($W$), length ($L$) and height ($H$).

Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations. 
<!-- TERMINOLOGY? rectangular solid -- impact --> top plane, front plane? Does the literture use "view"  -->
1.) if the aircraft were stationary, the area exposed to the debris impact would be the top view, and
2.) if the debris were stationary, it would be the front view.

Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes.
The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
At the same time, the movement of the re-entered debris object at regular en-route flight levels will be vertically [^balistic_descent].
This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as

$$
\varepsilon_k = \frac{s_{a} a_{h} + s_{d_k} a_{v}}{s_{d_k}}
$$

where 
* $s_{a}$ is the average aircraft speed, 
* $s_{d_k}$ is the average fall speed of the $k$-th debris object, 
* $a_T$ area of the aircraft as viewed from the top, and 
* $a_F$ area of the aircraft as viewed from the from the front of the aircraft.

@@

* constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$ 
  [[pag 8 @patera2008]]
* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, 
* aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$, 
  where $W_t$ is the wingspan, and $L_t$ the length for the specific aircraft
  type $t$
* aircraft area as viewed from the front, $a_F = W_t \times H_t = F_t$, 
  where $W_t$ is the wingspan, and $H_t$ the height for the specific aircraft
  type $t$

@@

$$
\varepsilon_t = \frac{C_{t} F_{t} + S A_{t}}{S}
$$

These aircraft properties are well-known. 
While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
The values $W_t$, $H_t$ and $L_t$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used \qty{1000}{m^2} for 
$\varepsilon_t$ in case of missing aircraft type.

@@

$$
p_{t,k}(h) = \varepsilon_t \rho_t(h) \sigma_k(h)
$$

[^balistic_descent]: *Even for a ballistic coefficient of \qty{50}{lbs/ft^2},
the flight path angle of a reentering object at \qty{40000}{ft} is 
\ang{-86.6}, which corresponds to nearly vertical descent.
Since most debris objects have ballistic coefficients equal to or 
less than \qty{50}{lbs/ft^2}, the vertical descent assumption is a very
accurate approximation.* [@patera2008]

@@

Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to re-entry even if initially on elliptical
ones.[^circular-orbit]

The angular position, $\theta$, of the re-entry point from the ascending node, 
see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
$\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in atmospheric drag
and the gravitational effects of the Equatorial bulge.
Hence we assume that the impact longitude, $\lambda$, is uniformely distributed.

The impact probability density, $\sigma(\varphi, i)$,
at latitude $\varphi$ of a reentring object whose orbit has inclination $i$
(see @fig-orbital-elements) is given by [[Equation 15 in @patera2008]]:

@@

where $R$ is the distance from the center of the Earth of the object potentially
being impacted by the debris.

So for example if a building is located at latitude 30 degrees and has an area 
of $\qty{1000}{m^2}$ and a piece of debris reenters the Earth’s atmosphere
from an orbit having an inclination of 45 degrees, the impact probability is 
given by

@@

where we used the equatorial value of $\qty{6378}{km}$, for the Earth's radius,
$R$, as the distance of the building from the center of the planet.

The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic
U shape.

@@

[^circular-orbit]: In fact drag at perigee will lower apogee and similarly drag
at apogee will lower perigee. But because drag at perigee is higher due to
denser atmosphere, the apogee  will decrease faster than perigee and the orbit
will then tend to be a circular one.

@@

If we consider historical uncontrolled debris reentry data, we can numerically
derive the debris impact probability as a weighting function by summing up and
averaging all single densities:

@@

Spatially combining $\sigma(\lambda,\varphi)$ for the H3 cells in the
\acr{NM} area we obtain the weightings per cell (per hour) as in
@fig-weighting-nm-area.

@@

Weightings for H3 cells in \acr{NM} area. 
:::

@@

When combining the hourly ($H$) aircraft density (per aircraft type) with the
relevant weightings per H3 cell, $h$, we can numerically compute the hourly
impact probability for each hexagonal cell as:

$$
 E_{h, H} =  \sum_t W(h) \cdot \rho_{t,H}(h) \cdot \varepsilon_t
$$

where 
* $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
* $W(h)$ is the median value of the weighting function over the area of hexagon
$h$,
* $\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
cell $h$ at hour $H$, and
* $\varepsilon_t$ is effective exposed area for aircraft of type $t$.

@@

# Application Use Case: KOSMOS 482 Descent Craft re-entry
<!-- I would make this a separate section in the paper -->

In few days (at the time of writing of this paper) at 6h May 10, 2025
($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
Earth's atmosphere [@kosmos482re-entry].
This is supposedly the lander module from a 1972 failed Soviet Venera mission [@SpaceReviewKosmos]
The module is reported to weigh about $\qty{480}{kg}$ and has a diameter of approximately $\qty{1}{m}$.
Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of reentry.

Once the reentry track is known, the H3 cells around \qty{35}{km} on each side 
can be identified to retrieve the values of collision expectancy.
It can then be decided whether it is advisable to close the airspace and
traffic around the area affected.

::: {#fig-kosmos}

```{r kosmos}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/example-image.png")
```

Casuality for KOSMOS 482 Descent Craft re-entry. 
:::

@@

We have calculated collision and casuality risk probabilities for the two
busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
as logically depicted in @fig-data-densities is available in the companion
GitHub repository of the paper at <https://github.com/espinielli/flight_density_space_debris>.
We have also developed a companion R package `aviodebris` to heavylift the
various data processing steps to produce the results in the paper;
it is available at <https://github.com/espinielli/aviodebris>.

@@

<!-- opening statement linking back to abstract/intro and movtivation for this work --------->
Space objects and debris re-entry is an increasing risk to aviation and the general public.
The higher pace of commercial space operations will see an increase of launches and subsequent re-entries.
Dependent on the mission, objects re-entering the atmosphere can take on differnt forms, shapes, and sizes.
On top, a higher number of these entries may be uncontrolled.
With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accomodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.

@@

The approach presented in this paper builds on previous research.
Our approach utilises H3-tesselation to reduce the spatial complexity.
This cell indexing approach reduces the computation significantly.

@@

As part of this research and to inform the decision-makers it is planned to explore the impact of megacostellations and flight growth forecast and what would happen in the mid- to long-term in varios scenarios of uncontrolled re-entries and the risk for European airspace.

@@

$$\sin(\varphi) = \sin(\theta)\sin(i)$$ {#eq-spheric-coords}

>>>>

crossref:
  eq-title: Equation     # (default is "")

@@

<!-- teaser opening --><!-- might need a rewrite - first jotting down -->
The quest for space is a driver for technological development.
Key milestones include the first human travel to space and circumnavigating the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
These developments helped to identify options for the wider use of space, benefits, and its commercialisation.
This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable production of goods in a non-gravity environment, services, and recently leisure.
The increased number of launches, however, also increased the number of artefacts <!-- falling down and --> re-entering the atmosphere.

@@

<!-- APPROACH / METHOD --------------------------------------------------- -->
<!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->
<!-- as a minimum, we should say something about the WHAT -->
In general, the method for estimating the risk to commercial aircraft from re-entering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
In that respect, a robust determination of flight density is a key aspect.

@@

This work focusses on the European region.
The hourly flight density estimation is based on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1].
To decrease spatial complexity, the data is binned using Uber's H3 geospatial indexing system [@uber-h3]. 
The reduction of the spatial complexity enables fast processing.
For this work, we employ a H3-cell-resolution of 3. 
This corresponds to hexagons of an average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

@@

H3 cells (red) at resolution 3 and their centres (blue) covering the bounding
box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded polygon).

@@

[^1]: NM coordinates \acr{ATFCM} operations on behalf of EUROCONTROL's 42 Member States and 2 Comprehensive Member States covering almost all \acr{ECAC} States, see blue polygon in @fig-cells.

<!-- next sentence feels like an orphan ... do we need this here? ------>
Next sections will develop the mathematical background for the evaluation of collision expectation of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
The chart shows the expected outcomes in terms of major traffic flows, and areas of high or low traffic.
For example, cells with a density of zero are clearly visible for areas of war or no-fly zones
(Ukraine, Libya or Syria). The major traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted showing densities of more than 10.

@@

```{r density-winter-res-3}
#| out-width: "110%"
#| width: "9cm"
#| echo: false
knitr::include_graphics("media/figures/flight_density_2024-07-05_3.png")
```

Flight density in the EUROCONTROL Area on the 5th of July 2024, the busiest day of the busiest week in 2024 (H3 cells at resolution $3$). The black hexagon in the South of UK marks the densest cell at 24.56 flights/hour, just around the airport of London Heathrow.
:::

@@

Safety and collision risk awareness is directly linked to the probability of a re-entering space object colliding with a traversing flight.
This probability, $p_k$, of a debris (point) object $k$, impacting an
aircraft (point) object over the area, $a$, is expressed as:

@@

where

* $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
* $\sigma(\vec{x})$ is the impact probability density of a debris object $k$
in the position $\vec{x}$, and 
* $\varepsilon_k$ is the effective exposed area, i.e., the effective risk area for the $k^{th}$ debris object impacting
the aircraft (see @sec-eea later on).

Our approach builds on the discretisation of the geospatial risk area.
For this purpose we use a hexagon-based tessellation pattern.
Uber's H3 method ensures non-overlapping hexagons (and 5 pentagons), with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
Using H3 to discretise the area of interest $a$ gives

@@

The impact probability $p_{k}(h)$ for each cell $h$ is determined by the following expression

$$
p_{k}(h) = \varepsilon_k \rho(h) \sigma_k(h).
$$ {#eq-impact-k-debris-cell}

@@

For the actual collision encounter between a re-entering object and an aircraft, we assume the aircraft impact area to be much larger than the debris object area.
Thus, the area spanned by the debris object is smaller than the aircraft dimensions. 
This assumption holds for practically all commercial aircraft within the scope of this study, and we can neglect the debris impact area.
Accordingly, we can also simplify the geometry of the aircraft and model it as a rectangular solid with dimensions of wingspan ($W$), length ($L$) and height ($H$).

Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations:
<!-- TERMINOLOGY? rectangular solid -- impact top plane, front plane? Does the literature use "view"  -->

1. if the aircraft were stationary, the area exposed to the debris impact would be the top face of the rectangular solid, and
2. if the debris were stationary, it would be the front face.

Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes.
The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
At the same time, the falling of the re-entering debris object at regular en-route flight levels will be vertical[^ballistic_descent].
This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as

$$
\varepsilon_k = \frac{s_{a} a_{h} + s_{d_k} a_{v}}{s_{d_k}}
$$ {#eq-effective-exposed-area}

where

* $s_{a}$ is the average aircraft speed, 
* $s_{d_k}$ is the average fall speed of the $k$-th debris object, 
* $a_T$ area of the aircraft as viewed from the top, and 
* $a_F$ area of the aircraft as viewed from the from the front of the aircraft.

@@

* constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$ 
  ([we adopt this value from pag 7-8 in @patera2008])
* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, under the
  simplification that the majority of the flight time spent cruising, 
* aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$, 
  where $W_t$ is the wingspan, and $L_t$ the length for the specific aircraft
  type $t$
* aircraft area as viewed from the front, $a_F = W_t \times H_t = F_t$, 
  where $W_t$ is the wingspan, and $H_t$ the height for the specific aircraft
  type $t$

@@

$$
\varepsilon_t = \frac{C_{t} F_{t} + S A_{t}}{S}
$$ {#eq-effective-exposed-area-simplified}

These aircraft properties are well-known. 
While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
The values $W_t$, $H_t$, $L_t$ and $C_{t}$ have been retrieved from [@performance_database;@skybrary;@doc8643]; we used $\varepsilon_t = \qty{1000}{m^2}$ missing entries.

@@

$$
p_{t,k}(h) = \varepsilon_t \rho_t(h) \sigma_k(h)
$$ {#eq-impact-k}

[^ballistic_descent]: *Even for a ballistic coefficient of \qty{50}{lbs/ft^2},
the flight path angle of a re-entering object at \qty{40000}{ft} is 
\ang{-86.6}, which corresponds to nearly vertical descent.
Since most debris objects have ballistic coefficients equal to or 
less than \qty{50}{lbs/ft^2}, the vertical descent assumption is a very
accurate approximation.* [@patera2008]

@@

Debris that re-enter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to re-entry even if initially on elliptical
ones.[^circular-orbit]

The angular position, $\theta = \omega + \nu$, of the re-entry point from the
ascending node, see @sec-orbitals, is uniformly distributed for uncontrolled
re-entries, $\sigma(\theta) = 1 / (2\pi)$.
This is due to uncertainties in atmospheric drag (among other things solar
activity can expand the atmosphere) and the gravitational effects of the
Equatorial bulge.
Hence we assume that the impact longitude, $\lambda$, is uniformly distributed.

The impact probability density, $\sigma(\varphi, i)$,
at latitude $\varphi$ of a re-entering object whose orbit has inclination $i$
(see @fig-orbital-elements) is given by ([Equation 15 in @patera2008]):

@@

where $R$ is the distance from the centre of the Earth of the object potentially
being impacted by the debris.

So for example if a building is located at latitude 30 degrees and has an area 
of $\qty{1000}{m^2}$ and a piece of debris re-enters the Earth’s atmosphere
from an orbit having an inclination of 45 degrees, the impact probability is 
given by

@@

where we used the equatorial value of $\qty{6378}{km}$, for the Earth's radius,
$R$, as the distance of the building from the centre of the planet.

The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic U shape with zero values for latitudes beyond $i$.
This density has been derived by simulating a full circular orbit at
\qty{550}{km} height with 10000  positions and calculating the relevant longitudes and latitudes at half degree resolution.

@@

[^circular-orbit]: In fact drag at perigee will lower apogee and similarly drag
at apogee will lower perigee. But because drag at perigee is higher due to
denser atmosphere (the perigee is closest to Earth), the apogee  will decrease faster than perigee and the orbit will then tend to become a circular one, see figure about apogee/perigee tracking in [@kosmos482reentry] for empirical evidence.

@@

If we consider historical uncontrolled debris re-entry data, we can numerically
derive the debris impact probability as a weighting function by summing up and
averaging all single densities:

@@

Spatially projecting $\sigma(\lambda,\varphi)$ on the H3 cells in the
\acr{NM} area we obtain the weightings per cell (per hour $H$), $W_H(h)$, as in
@fig-weighting-nm-area.

@@

Hourly weightings for H3 cells, $W_H(h)$, in \acr{NM} area. 
:::

Then @eq-impact-k becomes

$$
p_t(h) = \varepsilon_t \rho_t(h) \sigma(h).
$$ {#eq-impact-cell}

@@

When combining the hourly ($H$) aircraft density (per aircraft type) with the
relevant weightings per $h$ cell, $W_H(h)$, we can numerically compute the
hourly impact probability for each hexagonal cell as:

$$
 E_{h, H} =  \sum_t W_H(h) \cdot \rho_{t,H}(h) \cdot \varepsilon_t
$$

where

* $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
* $W(h)$ is the median value of the weighting function over the area of hexagon
$h$,
* $\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
cell $h$ at hour $H$, and
* $\varepsilon_t$ is effective exposed area for aircraft of type $t$.

@@

# Use Case: KOSMOS 482 Descent Craft re-entry
<!-- I would make this a separate section in the paper -->

In few days (at the time of writing of this paper) at 23:06 May 10, 2025
($\pm 1.8$ days), the KOSMOS 482 Descent Craft should re-enter the
Earth's atmosphere [@kosmos482reentry].
Supposedly the lander module from a 1972 failed USSR Venera mission [@SpaceReviewKosmos],
it is reported to weigh about $\qty{480}{kg}$ and to have a diameter of approximately $\qty{1}{m}$.
Being built to survive Venus's atmosphere which is much denser and hotter than Earth's one (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of re-entry.
The densest cell, `83194afffffffff` (area \qty{11131585027}{m^2}), at 24.56 flights/hour has centroid latitude
of \ang{50.94369} so within the latitude range of $[\ang{-51.7}, \ang{51.7}]$.
Using an average value for effective exposed area of \qty{1000}{m^2},
we can quickly calculate the collision expectation as

$$
\begin{aligned}
E &= w_h \cdot \sigma_h \cdot \bar{\varepsilon} / area_h \\
  &= 10^{-6} \cdot 24.56 \cdot 1000 / 11131585027 = 2.2\cdot 10^{-12}
\end{aligned}
$$

The risk is very low.

Eventually once the re-entry track is known, the H3 cells around \qty{35}{km} on
each side  can be identified to retrieve the values of collision expectancy.
It can then be decided whether it is advisable to close the airspace and re-route
traffic around the area affected.

@@

We have calculated collision and casualty risk probabilities for the two
busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
as logically depicted in @fig-data-densities is available in the companion
GitHub repository of the paper at <https://github.com/euctrl-pru/flight_density_space_debris>.
We have also developed a companion R package `aviodebris` to heavy-lift the
various data processing steps to produce the results in the paper;
it is available at <https://github.com/euctrl-pru/aviodebris>.

@@

<!-- opening statement linking back to abstract/intro and motivation for this work --------->
Space objects and debris re-entry is an increasing risk to aviation and the general public.
The higher pace of commercial space operations will see an increase of launches and subsequent re-entries.
Dependent on the mission, objects re-entering the atmosphere can take on different forms, shapes, and sizes.
On top, a higher number of these entries may be uncontrolled.
With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accommodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.

@@

The approach presented in this paper builds on previous research.
Our approach utilises H3-tessellation to reduce the spatial complexity.
This cell indexing approach reduces the computation significantly.

@@

As part of this research and to inform the decision-makers it is planned to explore the impact of megaconstellations and flight growth forecast and what would happen in the mid- to long-term in various scenarios of uncontrolled re-entries and the risk for European airspace.

@@

$$
\sin(\varphi) = \sin(\theta)\sin(i)
$$ {#eq-spheric-coords}
